Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-09-03 Manuscript/manuscript.Rmd
discussion edits

A key factor in locating TAD boundaries is Hi-C data "resolution". Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Lower resolutions contribute to larger bin sizes, which lead to variability in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to define sub-domains and the boundaries that demarcate them. Furthermore, even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-Sequencing experiments used to annotate the genome, which ranges in the tens of base pairs. This resolution gap between 1D and 3D genomic maps can make it difficult to disentangle the role of genomic factors related to TAD boundaries, thereby limiting our understanding of epigenomic regulation and its potential link to chromatin architecture. We hypothesize that, by leveraging genomic annotations--defined by high resolution ChIP-Seq data--known to be enriched at TAD boundaries, we may be able more precisely predict the locations of TAD boundaries at higher resolutions.

@@

ChIP-seq assays offer much higher resolutions (often in the tens of bases) compared to Hi-C sequenced data, which can be in the hundreds of kilobases. This disparity in resolution can affect a TAD caller's ability to precisely locate TADs.[??? This is for the introduction, justifying motives for your work]

@@

# DISCUSSION

In this study we developed an ensemble framework to systematically compare model performances for different combinations of re-sampling and feature engineering techniques aimed at determining a consensus approach to predicting TAD boundary formation using genomic annotations. We provide direct comparisons of cell line specific random forest models built on a variety of different data dependent characteristics, including 4 contact matrix resolutions ($\{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$), 3 predictor-types ($\{\textit{OC}, \textit{OP}, \textit{Distance}\}$), and 4 re-sampling techniques ($\{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$), for both the GM12878 and K562 cell lines. [??? Learn Markdown] [??? What's the summary of your work?]

The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Using the framework we demonstrated that random forest models built on imbalanced data exhibited poor performance, regardless of cell line, resolution, or predictor type. In our framework, [??? Too much "framework" mentioning] we opted to leave the test set as a completely untouched holdout set, only performing the re-sampling, and subsequent learning, on the training set. This was done so as to incur the least amount of bias and so that all conclusions could be made soley between model characteristics. As a result, we showed that Matthews Correlation Coefficient was much more sensitive to the imbalance present in the test set, compared to more popular metrics like AUROC, which attain artificially high values by correctly predicting majority class samples [??? I don't see results for it]. When comparing re-sampling techniques, we found that SMOTE re-sampling outperformed models built on both simple ROS or RUS, implying that, in fact, a combination of both random under-sampling and random over-sampling is preferred in TAD boundary prediction. Furthermore, the most improvement [??? In what?] was seen when combining SMOTE resampling on distance-type features when building random forest classifiers. These two characteristics exhibited consistently better performance results across both the GM12878 and K562 cell lines. This confirmed our hypothesis that the most granular and spatially informative representation of the relationship between TAD boundaries and annotations resulted in the best performing model.

When applying SMOTE re-sampling, with distance-type predictors, on transcription factor binding sites alone, we found that random forest classifiers performed better when compared to other annotation-specific models, including histone modifications and chromatin state segmentations. Likewise the performance of TFBS specific models was comparable to models with all annotations combined, across both cell line and resolution. This suggests that the majority of the key molecular drivers aiding in the formation of TAD boundaryies are transctiption factors, and that modifications at histone tails such as acetylation or methylation play less of a role.

We also observed that, while the best set of data characteristics were consistent across resolutions, performances of classifiers built on said characteristics tended to decline as resolution decreased. For both cell lines, top performance was achieved at the 10 kb data resolution, while the worst performance was obtained at the 50 kb and 100 kb resolution for the GM12878 and K562 cell line respectively. This suggests that TAD boundary prediction is intimately related to the resolution that boundaries are called on. This result can be attributed to one of two factors in the scheme of our framework. First, at higher resolutions, by model construction genomic bins were smaller, thereby resulting in more accurate profiles for the genomic annotations [?? Don't understand this sentence ??]. Secondly, because we are mainly concerned with TAD boundaries, and not the hierarchical structure of domains as a whole, the sparsity of contact matrices at higher resolutions does not impair boundary prediction [?? Doesn't this contradict your results about low resolution causing worse boundary prediction ??]. Conversely, lower resolutions imply larger bin windows and lead to less specificity in boundary location on the linear genome, resulting in worse performances. Given the trend in performances across resolution, future research is necessary in determining the optimal resolution to use when predicting TAD boundaries [?? Might want to qualify the statement in results about 10kb data being best by saying that it only applies to your data ??]. [??? This paragraph is a mess]

Given the formulation of our framework, we were not able to make direct comparisons between cell lines. This was due to the fact that we opted to use cell line specific annotations, and there existed some annotations that were available on the K562 cell line that were not available to GM12878, and vice versa. [??? Did they provide additional advantage?] However, we were able to confirm that final models built on the optimal set of data characteristics identified consistent and well known proteins related to the 3D structure of chromatin, for both cell lines. Our results indicated that the zinc finger protein CTCF, as well components of the cohesin complex, SMC3 and RAD21, were found to be the key molecular drivers most predictive of TAD boundary formation, consistent across cell line and resolution. In addition, models derived from our framework recognized known mediaters of long range enhancer-promoter interactions, the ZNF143 and YY1 transcription factors, as being highly predictive of TAD boundaries.

In summary, we have shown that TAD boundary prediction is a multi-faceted problem composed of many data irregularities and characteristics that previous model have failed to address. We have shown that our proposed framework suitably addresses each of these components, while also providing the optimal considerations to make in an effort to build consensus classifiers used to predict TAD boundary formation [?? formation or location. Formation implies some temporal element ??] from functional genomic annotations. Given the stability of our results, this opens up the possibility of widening our framework to include a broader set of classifiers, cell lines, and Hi-C contact matix resolutions that will inevitably assist us in gaining a better understanding of the mechanisms associated with the architecture of the human genome.

>>>>

A key factor in locating TAD boundaries is Hi-C data "resolution". Resolution determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Lower resolutions contribute to larger bin sizes, which lead to variability in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define sub-domains and the boundaries that demarcate them. Furthermore, even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-Sequencing experiments used to annotate the genome, which ranges in the tens of base pairs. This resolution gap between 1D and 3D genomic maps can make it difficult to disentangle the role of genomic factors related to TAD boundaries, thereby limiting our understanding of epigenomic regulation and its potential link to chromatin architecture. We hypothesize that, by leveraging genomic annotations--defined by high resolution ChIP-Seq data--known to be enriched at TAD boundaries, we may be able more to precisely predict the locations of TAD boundaries at higher resolutions.

@@

@@

# Discussion

In this study we developed an ensemble framework to systematically compare model performances for different combinations of Hi-C resolution, data re-sampling, and feature engineering techniques. By implementing our framework across multiple cell lines, we showed that there exists a consensus approach toward precise TAD boundary prediction. We showed that random forest classification models trained on domain data extracted from 10 kb resolution maps, using random over- and under-sampling to account for class imbalance, with a distance-type predictor space yield optimal model performances for both the GM12878 and K562 cell lines.

Previous studies tended to focus only on specific sets of genomic annotations including chromatin states, histone modifications, or transcription factor binding sites alone. Using our framework, we were able to conclude that simply building random forest classifiers on cell line specific transcription factor binding sites alone can yield near optimal model performance, thereby significantly reducing the feature space and computational complexity of the model. Specifically, we found that the CCCTC-binding factor (CTCF) and cohesin complexes (RAD21 and SMC3) were the most important features of TAD boundary prediction, confirming their enrichment at TAD boundaries as part of the loop extrusion model [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, we found the the zinc-finger protein ZNF143 was also given high predictive importance. This confirmed the known role of ZNF143 as a chromatin-looping factor that contributes to the architectural foundation of the genome by providing sequence specificity at promoters connected with distal regulatory elements [@bailey2015znf143;@gonzalez2017znf143].
